Informing policy via dynamic models: Eliminating cholera in Haiti

### 2022-11-30 ms.Rnw
minor edits

# Load Required packages
library(knitr)
library(pomp)
library(panelPomp)
library(spatPomp)
library(foreach)
library(doParallel)
library(doRNG)
library(haitipkg)
library(tidyverse)

@@

The four independent teams were given the task of estimating the
potential effect of prospective oral cholera vaccine (OCV) programs.
While OCV is accepted as a safe and effective tool for controlling the spread of cholera, the global stockpile of OCV doses remains limited \citep{pezzoli20}.
Advances in OCV technology and vaccine availability, however, raised the possibility of planning a national vaccination program \citep{lee20}.
In the study, certain data were shared between the groups, including demography and vaccination history; vaccine efficacy was also fixed at a shared value between groups.
Beyond this, the groups made autonomous decisions on what to include and exclude from their models.
The groups largely adhered to existing guidelines on creating models to inform policy \citep{behrend20,saltelli20} and, despite their autonomy, obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
Although several cholera chases have recently been detected in Haiti (\citep{TODO}),
their conclusion is inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 \citep{trevisin22} and \jwc{TODO}.

@@

\jwc{TODO: CHECK} It should be universal practice to present measures of goodness of fit for published models, and mechanistic models should be compared against benchmarks.
In our literature review of the Haiti cholera epidemic, no benchmark models were considered in any of the 24 papers which calibrated a mechanistic model to data in order to obtain scientific conclusions.
Including benchmarks would help authors and readers to detect and confront any major statistical limitations of the proposed mechanistic models.
In addition, the published goodness of fit provides a concrete point of comparison for subsequent scientific investigations.
When combined with online availability of data and code, objective measures of fit provide a powerful tool to accelerate scientific progress, following the paradigm of the {\it common task framework} \citep[][Sec.~6]{donoho17}.

>>>>

# Load Required packages
library(knitr)
library(pomp)
library(spatPomp)
library(foreach)
library(doParallel)
library(doRNG)
library(haitipkg)
library(tidyverse)

@@

The four independent teams were given the task of estimating the
potential effect of prospective oral cholera vaccine (OCV) programs.
While OCV is accepted as a safe and effective tool for controlling the spread of cholera, the global stockpile of OCV doses remains limited \citep{pezzoli20}.
Advances in OCV technology and vaccine availability, however, raised the possibility of planning a national vaccination program \citep{lee20}.
In the study, certain data were shared between the groups, including demography and vaccination history; vaccine efficacy was also fixed at a shared value between groups.
Beyond this, the groups made autonomous decisions on what to include and exclude from their models.
The groups largely adhered to existing guidelines on creating models to inform policy \citep{behrend20,saltelli20} and, despite their autonomy, obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
Although several cholera cases have recently been detected in Haiti (\citep{world2010outbreak}),
their conclusion is inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 \citep{trevisin22} and October, 2022.

@@

It should be universal practice to present measures of goodness of fit for published models, and mechanistic models should be compared against benchmarks.
In our literature review of the Haiti cholera epidemic, no benchmark models were considered in any of the 32 papers that used dynamic models to describe the cholera in order to obtain scientific conclusions.
Including benchmarks would help authors and readers to detect and confront any major statistical limitations of the proposed mechanistic models.
In addition, the published goodness of fit provides a concrete point of comparison for subsequent scientific investigations.
When combined with online availability of data and code, objective measures of fit provide a powerful tool to accelerate scientific progress, following the paradigm of the {\it common task framework} \citep[][Sec.~6]{donoho17}.
